Learning communication systems through recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behavior, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each object.\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.

>>>>

\subsection{Paradoxes in optimal models of pragmatic learning}
% When modelling human behavior, it's useful to begin with optimal
% models; this has the advantage that when we are then forced to
% deviate from the purely rational approach, it's generally in a
% theoretically interpretable way. However,
Multi-agent interactions are difficult to model in a normative or
optimal framework without falling prey to paradox. For example,
consider a simple model of the agents in the above game. First we
define a \textit{literal listener} $L_0$. This agent has a
\textit{lexicon} which probabilistically maps words to meanings;
specifically, it assigns each word $w$ a vector describing the extent
to which this word provides evidence for each possible object\footnote{We
  assume words refer directly to objects, rather than to abstract
  semantic features. Our simplification is without loss of
  generalization, however, because we can interpret our model as
  marginalizing over such a representation, with our literal
  $P_\lex(\obj | \word) = \sum_\text{features} P(\obj |
  \text{features}) P_\lex(\text{features} | \word)$.} Each entry in
the vector is in $(0, 1)$, and together they sum to 1:
\begin{align*}
P_{L_0}(\obj | \word, \lex) &\propto \lex(\word, \obj) \times \prior.
\end{align*}
Confronted with such a listener, a probabilistically rational speaker
should attempt to choose a word which soft-maximizes the probability
that the listener will assign to the target object---modulated by the
effort or cost associated with producing this word:
\begin{align*}
P_{S_1}(\word | \obj, \lex) &\propto \exp\Big(\lambda \big(\log P_{L_0}(\obj | \word, \lex) - \text{cost}(\word)\big)\Big).
\end{align*}
But given this speaker, then the naive $L_0$ strategy is not
optimal. Instead, listeners should use Bayes rule to invert the
speaker's decision procedure \cite{frank2012}:
\begin{align*}
P_{L_2}(\obj | \word, \lex) &\propto P_{S_1}(\word | \obj, \lex) \times \prior.
\end{align*}
Now the paradox becomes apparent. Given such a listener, it is no
longer optimal for speakers to implement strategy $S_1$; instead, they
should implement strategy $S_3$ which soft-maximizes $P_{L_2}$ instead
of $P_{L_0}$. And then listeners ought to implement $L_4$, and so on.
